What we {\sf eval} in the shadows

### 2021-03-31 paper/main.tex
wip

%\documentclass[USenglish,cleveref, autoref, thm-restate]{lipics-v2019}
\documentclass[acmsmall]{acmart}
\usepackage{listings,hyperref,multirow,paralist,xspace,url,wrapfig}
\usepackage{tabularx}
\input{macro}
\newcommand{\mypara}[1]{\medskip\noindent\emph{#1}\xspace}
\newcommand{\NOTE}[1]{{\it Note: #1}\xspace}

@@

\title{Why do we Eval?\\\Large A large-scale study of Eval usage in R}
%\titlerunning{Why Eval (in R)?}

@@

\begin{lstlisting}
 eval
   <- function(expr,
                envir = parent.frame(),
                enclos = if(is.list(envir)) parent.frame() else baseenv())
          { ...body... }
\end{lstlisting}
The parameters are \c{expr}, the value to be evaluated, \c{envir}, the
environment in which evaluation happens, and \c{enclos}, the
environment to look up objects not found in \c{envir} (by default
either the top level, \c{baseenv} or the environment of the caller).
The \c{expr} parameter can take values of different type, for our
purposes we focus on the most common, {\tt expression}s. It is easiest
to think of an \c{expression} as an abstract syntax tree as returned
by R's parser. The simplest way to create an \c{expression} is to call
\c{quote} passing some R expression:
\begin{lstlisting}
  > quote(a + b)
  # a + b
\end{lstlisting}
Here the return value is an abstract syntax tree. From a
string, use  the \c{parse} function:
\begin{lstlisting}
  > parse(text="a+b")
  # expression(a+b)
\end{lstlisting}
The most common way to create an expression is to extract it from an
argument. Each argument comes packaged inside a promise which retains
the source code of the argument. Consider the following function
definition, \c{f} is a function with a single parameter \c{x}. When
the call \c{f(a+b)} is evaluated, \c{a+b} is stored inside a newly
created promise.
\begin{lstlisting}
  > f <- function(x) substitute(x), list(a=1)))
  > f(a+b)
  # expression(1+b)
  > substitute(a+b,list(a=1,b=2))
  # 1+2
\end{lstlisting}
The call to \c{substitute(x)} extracts, from the promise bound to
\c{x}, the expression passed in the call to \c{f}, i.e., \c{a+b}. The
\c{substitute(exp,env)} function takes two arguments, the second is an
object that can be used as an environment. This can be either a list
of named values, a data frame or an actual environment. \c{substitute}
will look for occurrences of each symbol from \c{env} in \c{exp} and
replace them with their value taken from \c{env}.
\begin{lstlisting}
  > environment()
  # <environment: R_GlobalEnv>
\end{lstlisting}
The above shows how to access the current environment, this can belong
to the current function or, as above, the top level. Environments are
nested, with each environment having a parent. This nesting is used
when looking up names. When calling \c{new.env}, the default parent is
the current environment. The chain of environments can be traversed
with \c{parent.env}, ultimately one arrives at \c{emptyenv}.
\c{parent.frame} and \c{sys.frame} grant access to environments
further up in the calling stack, up to the global environment
\c{.GlobalEnv}, and packages. One can also directly read, modify or
create new bindings, given any environment:
\begin{itemize}
\item \c{env$v} and \c{get("v",envir=env)} read variable \c{v} in
  environment \c{env};
\item \c{env$v<-2} and \c{assign("v",2,envir=env)} write 2 in
  \c{v}. If not found, \c{v} is created in \c{env}.
\end{itemize}
\noindent Environments are often used as hashtables by programmers as
they have reference semantics (other are copy-on-write) and have a
built-in string look up. across function calls, or to create package
namespaces.
\begin{lstlisting}
  evalq <- function(exp,env,enclos) eval(quote(exp),env,enclos)
  eval.parent <- function(exp,n) eval(exp,parent.frame(n))
  local <- function(exp,env,enclos) evalq(exp,new.env())
\end{lstlisting}
The three \eval variants can be expressed as calls to \eval. The
\c{evalq} form quotes the argument to prevent it from being evaluated
in the current environment. The \c{eval.parent(e,n)} form evaluates
\c{e} in the environment of the {\tt n}-th caller of this function.
Finally, \c{local} evaluates an \c{exp} in a new environment to
avoid polluting the current one.

@@

\mypara{Limitations.}
%
% We turn off the bytecode compiler for this study. The bytecode
% compiler can also call \eval.
%
We do not get source locations for \CranUndefinedRnd \eval calls. In
these cases \eval is either passed as an argument to a higher-order functions or
is defined in a function returned by a higher-order function, or it can be a
generated called from package native code. However, this is a meager
\CranUndefinedRatio of all \eval calls and is unlikely to affect our
analysis. We ignore calls to the native \eval function exposed by R. We also
ignore the \c{rlang::tidy_eval} function which uses native \eval internally
because \c{rlang} is used to implement a DSL for data analysis in R. It
introduces a new first-class promise object called \c{quosure} for which it
implements special evaluation support in \c{tidy_eval}.

@@

Arguments to \eval can be either code or values, in the latter case
\eval has no effect. Code accounts for \Codepercent of arguments in
our corpus. More specifically, \Symbolpercent of arguments are
\c{symbol}s (i.e. these are single variables such as \c{x}),
\Languagepercent are \c{language} object (an internal R type roughly
corresponding to a single expression, such as \c{x+1}), and
\Expressionpercent are \c{expression} objects (multiple statements).
Further inspection reveals that most symbols, \Ggplotsymbolpercent to
be exact, come from a single call in the \c{ggplot2} package and have
have the value \c{_inherit}. Even if we remove this outlier, the
majority of \evals are simple expressions.

@@

We looked at the environments passed to all the \eval calls in our
corpus. Table~\ref{tab:environments} summarizes the results. A numeric
environment class \c{n} denotes the environment of the $n$-th call
stack frame from the current function. \c{global} denotes the
top-level environment and \c{list} denotes a list passed for
evaluation of formulas. Environment classes of the form $n+$ denote
the $n$-th environment extended with a new environment. The new
environment provides a limited form of sandboxing. All assignments
using the \c{\<-} function occur inside it and prevent the extended
environment from mutation. However, it is still possible to mutate the
extended environment using the \c{\<\<-} or \c{assign} functions; but,
that happens rarely.

@@

The \c{imchange} function of package \c{imager} makes it possible to
modify images using a dedicated formula syntax using
\c{\~}.\footnote{Inspired by {map} in package \emph{purr}.} Here,
\eval is evaluated in \c{newenv}, which creates a new environment that
inherits from \c{parent.frame()} by default (classified as 1+).

@@

\begin{table}[!h]  \centering
\begin{tabular}{@{}l@{~}|@{~}r@{~}r@{}} \hline
Expression & Eval Call &  \% \\\hline
\c{Environment} &                                  989302   & 61\%\\
\c{column[rows] <<- what} &                        55677    & 3.5\%\\
\c{function(value) freduce(value, `_function_list`)} & 37251& 2.3\%\\
\c{NULL} &                         32005    & 2\%\\
\c{List} &                         22293    & 1.4\%\\
\c{c("default", "default2012", "default2011" ...}& 20610    & 1.3\%\\
\c{force(..1)}            &                        20461    & 1.3\%\\
\c{alist(`_spec`)}       &                         18532    & 1.2\%\\
\c{inner}               &                          18530    & 1.2\%\\
\c{String Vector}      &                           17487     & 1.1\%\\
\end{tabular}\caption{Top ten eval calls in CRAN} \label{C}
\end{table}

The expression \c{Environment} occurs because of the four callsites
explained above, \c{ggplot2::ggproto}, \c{R6::generator_funs},
\c{future::backtrace} and \c{RModel::str.RMmodel}. The next
expression, \c{column[rows] <<- what}, is used inside the
\c{plyr::rbind.fill} function to merge data frames by assigning
concatenated vectors to rows. The \c{<<-} operator is interesting in
that it skips the current scope and assigns in a parent scope in which
the variable is already present. In our corpus, all these \eval calls
contribute to a single side-effect. The expression \c{function(value)
  freduce(value, `_function_list`)} arises from the
\c{magrittr::\%>\%} function which is a pipe operator that pipes the
output of previous command to the next one. The expression is
evaluated in a custom environment to create a function binding for
evaluating the components of the pipe.\c{String Vector} and \c{List}
also arise from the same function when a string or a list is piped
using the \c{\%>\%} function into the next expression. The \c{NULL}
arises from \c{R6::generator_funs} function when the \eval is passed a
\c{NULL} argument by the \c{DataMask_generator} package. The
\c{c("default", "default2012", "default2011" ...} pattern arises from
\c{copula::polyG} where it reflectively access the default expression
for its formal parameter and evaluates it. The \c{force(..1)} and
\c{alist(`_spec`)} patterns occur in \c{glue::glue_data} function
which concatenates and interpolates strings. The two patterns occur
because the function captures unevaluated unnamed arguments and maps
the evaluation of \c{force(..1)} on them. The \c{force} function
forces promises and returns the result of evaluation. The \c{inner}
pattern arises from \c{glue::identity_transformer} which enables the
creation of custom transformation functions for affecting the
interpolation and concatenation of input by the \c{glue} package.

>>>>

\documentclass[screen,acmsmall]{acmart}
\settopmatter{printfolios=true,printccs=true,printacmref=true}

\bibliographystyle{ACM-Reference-Format}
\citestyle{acmauthoryear}   %% For author/year citations
\usepackage{listings,hyperref,multirow,paralist,xspace,url,wrapfig}
\usepackage{tabularx}
\input{macro}
\newcommand{\mypara}[1]{\medskip\noindent\emph{#1}\xspace}
\newcommand{\NOTE}[1]{{\it Note: #1}\xspace}

@@

\title{What we {\sf eval} in the shadows}
\subtitle{A large-scale study of {\sf eval} usage in R programs}

%% Aviral Goel
%% Pierre Donat-Bouillud
%% Filip Krikava
%% Jan Vitek

%% What is the right order?

@@

\begin{lstlisting}
 eval <- function(expr, envir = parent.frame(),
           enclos = if(is.list(envir)) parent.frame() else baseenv()) {...body...}
\end{lstlisting}
The parameters are \c{expr}, the value to be evaluated, \c{envir}, the
environment in which evaluation happens, and \c{enclos}, the
environment to look up objects not found in \c{envir} (by default
either the top level, \c{baseenv} or the environment of the caller).
The \c{expr} parameter can take values of different type, for our
purposes we focus on the most common, {\tt expression}s. It is easiest
to think of an \c{expression} as an abstract syntax tree as returned
by R's parser. The simplest way to create an \c{expression} is to call
\c{quote} passing some R expression:
\begin{lstlisting}
  > quote(a + b)
  # a + b
\end{lstlisting}
Here the return value is an abstract syntax tree. From a
string, use  the \c{parse} function:
\begin{lstlisting}
  > parse(text="a+b")
  # expression(a+b)
\end{lstlisting}
The most common way to create an expression is to extract it from an
argument. Each argument comes packaged inside a promise which retains
the source code of the argument. Consider the following function
definition, \c{f} is a function with a single parameter \c{x}. When
the call \c{f(a+b)} is evaluated, \c{a+b} is stored inside a newly
created promise.
\begin{lstlisting}
  > f <- function(x) substitute(x), list(a=1)))
  > f(a+b)
  # expression(1+b)
  > substitute(a+b,list(a=1,b=2))
  # 1+2
\end{lstlisting}
The call to \c{substitute(x)} extracts, from the promise bound to
\c{x}, the expression passed in the call to \c{f}, i.e., \c{a+b}. The
\c{substitute(exp,env)} function takes two arguments, the second is an
object that can be used as an environment. This can be either a list
of named values, a data frame or an actual environment. \c{substitute}
will look for occurrences of each symbol from \c{env} in \c{exp} and
replace them with their value taken from \c{env}.
\begin{lstlisting}
  > environment()
  # <environment: R_GlobalEnv>
\end{lstlisting}
The above shows how to access the current environment, this can belong
to the current function or, as above, the top level. Environments are
nested, with each environment having a parent. This nesting is used
when looking up names. When calling \c{new.env}, the default parent is
the current environment. The chain of environments can be traversed
with \c{parent.env}, ultimately one arrives at \c{emptyenv}.
\c{parent.frame} and \c{sys.frame} grant access to environments
further up in the calling stack, up to the global environment
\c{.GlobalEnv}, and packages. One can also directly read, modify or
create new bindings, given any environment:
\begin{itemize}
\item \c{env\$v} and \c{get("v",envir=env)} read variable \c{v} in
  environment \c{env};
\item \c{env\$v<-2} and \c{assign("v",2,envir=env)} write 2 in
  \c{v}. If not found, \c{v} is created in \c{env}.
\end{itemize}
\noindent Environments are often used as hashtables by programmers as
they have reference semantics (other are copy-on-write) and have a
built-in string look up. across function calls, or to create package
namespaces.
\begin{lstlisting}
  evalq <- function(exp,env,enclos) eval(quote(exp),env,enclos)
  eval.parent <- function(exp,n) eval(exp,parent.frame(n))
  local <- function(exp,env,enclos) evalq(exp,new.env())
\end{lstlisting}
The three \eval variants can be expressed as calls to \eval. The
\c{evalq} form quotes the argument to prevent it from being evaluated
in the current environment. The \c{eval.parent(e,n)} form evaluates
\c{e} in the environment of the {\tt n}-th caller of this function.
Finally, \c{local} evaluates an \c{exp} in a new environment to
avoid polluting the current one.

@@

\mypara{Limitations.}
%
% We turn off the bytecode compiler for this study. The bytecode
% compiler can also call \eval.
%
We do not get source locations for \CranUndefinedRnd \eval calls. In
these cases \eval is either passed as an argument to a higher-order functions or
is defined in a function returned by a higher-order function, or it can be a
generated called from package native code. However, this is a meager
\CranUndefinedRatio of all \eval calls and is unlikely to affect our
analysis. We ignore calls to the native \eval function exposed by R. We also
ignore the \c{rlang::tidy\_eval} function which uses native \eval internally
because \c{rlang} is used to implement a DSL for data analysis in R. It
introduces a new first-class promise object called \c{quosure} for which it
implements special evaluation support in \c{tidy\_eval}.

@@

Arguments to \eval can be either code or values, in the latter case
\eval has no effect. Code accounts for \Codepercent of arguments in
our corpus. More specifically, \Symbolpercent of arguments are
\c{symbol}s (i.e. these are single variables such as \c{x}),
\Languagepercent are \c{language} object (an internal R type roughly
corresponding to a single expression, such as \c{x+1}), and
\Expressionpercent are \c{expression} objects (multiple statements).
Further inspection reveals that most symbols, \Ggplotsymbolpercent to
be exact, come from a single call in the \c{ggplot2} package and have
have the value \c{\_inherit}. Even if we remove this outlier, the
majority of \evals are simple expressions.

@@

We looked at the environments passed to all the \eval calls in our corpus.
Table~\ref{tab:environments} summarizes the results. A numeric environment class
\c{n} denotes the environment of the $n$-th call stack frame from the current
function. \c{global} denotes the top-level environment and \c{list} denotes a
list passed for evaluation of formulas. Environment classes of the form $n+$
denote the $n$-th environment extended with a new environment. The new
environment provides a limited form of sandboxing. All assignments using the
$\leftarrow$ function occur inside it and prevent the extended environment from
mutation. However, it is still possible to mutate the extended environment using
the $\leftarrow$ or \c{assign} functions; but, that happens rarely.

@@

The \c{imchange} function of package \c{imager} makes it possible to
modify images using a dedicated formula syntax using
\~.\footnote{Inspired by {map} in package \emph{purr}.} Here,
\eval is evaluated in \c{newenv}, which creates a new environment that
inherits from \c{parent.frame()} by default (classified as 1+).

@@

\begin{table}[!h]  \centering
\begin{tabular}{@{}l@{~}|@{~}r@{~}r@{}} \hline
Expression & Eval Call &  \% \\\hline
\c{Environment} &                                  989302   & 61\%\\
\c{column[rows] <<- what} &                        55677    & 3.5\%\\
\c{function(value) freduce(value, `function.list`)} & 37251& 2.3\%\\
\c{NULL} &                         32005    & 2\%\\
\c{List} &                         22293    & 1.4\%\\
\c{c("default", "default2012", "default2011" ...}& 20610    & 1.3\%\\
\c{force(..1)}            &                        20461    & 1.3\%\\
\c{alist(`spec`)}       &                         18532    & 1.2\%\\
\c{inner}               &                          18530    & 1.2\%\\
\c{String Vector}      &                           17487     & 1.1\%\\
\end{tabular}\caption{Top ten eval calls in CRAN} \label{C}
\end{table}

The expression \c{Environment} occurs because of the four callsites
explained above, \c{ggplot2::ggproto}, \c{R6::generator\_funs},
\c{future::backtrace} and \c{RModel::str.RMmodel}. The next
expression, \c{column[rows] <<- what}, is used inside the
\c{plyr::rbind.fill} function to merge data frames by assigning
concatenated vectors to rows. The \c{<<-} operator is interesting in
that it skips the current scope and assigns in a parent scope in which
the variable is already present. In our corpus, all these \eval calls
contribute to a single side-effect. The expression \c{function(value)
  freduce(value, `function.list`)} arises from the
\c{magrittr::\%>\%} function which is a pipe operator that pipes the
output of previous command to the next one. The expression is
evaluated in a custom environment to create a function binding for
evaluating the components of the pipe.\c{String Vector} and \c{List}
also arise from the same function when a string or a list is piped
using the \c{\%>\%} function into the next expression. The \c{NULL}
arises from \c{R6::generator\_funs} function when the \eval is passed a
\c{NULL} argument by the \c{DataMask\_generator} package. The
\c{c("default", "default2012", "default2011" ...} pattern arises from
\c{copula::polyG} where it reflectively access the default expression
for its formal parameter and evaluates it. The \c{force(..1)} and
\c{alist(`spec`)} patterns occur in \c{glue::glue\_data} function
which concatenates and interpolates strings. The two patterns occur
because the function captures unevaluated unnamed arguments and maps
the evaluation of \c{force(..1)} on them. The \c{force} function
forces promises and returns the result of evaluation. The \c{inner}
pattern arises from \c{glue::identity\_transformer} which enables the
creation of custom transformation functions for affecting the
interpolation and concatenation of input by the \c{glue} package.
